---
title: "Multi-trait SuSiE"
author: "Yongjin Park"
---

How can we distinguish shared effects across multiple traits from trait-specific effects?  Inspired by [`susieR`](https://github.com/stephenslab/susieR), we extend the integrative single-effect estimation steps to a multi-trait problem. 

### A toy example

Let's generate a toy example to illustrate the problem.

```{r}
set.seed(1331)

.rnorm <- function(d1, d2) matrix(rnorm(d1*d2), d1, d2)

n <- 200                                  # sample size
p <- 1000                                 # number of predictors/variants
m <- 10                                   # traits/outcomes
h2 <- .2                                  # heritability/proportion of variance exampled by X
X <- .rnorm(n, p)                         # this can be replaced with a real genotype matrix
common <- sample(p, 1)                    # common effect across all the traits
.theta <- .rnorm(length(common), m)       # the effect size may fluctuate
y <- X[, common, drop = FALSE] %*% .theta # 

other <- c()
for(k in 1:m){
    set.k <- sample(p, 1)                 # sample trait-specific ones
    other <- c(other, set.k)              #
    y[, k] <- y[, k] + X[, set.k, drop=FALSE] %*% .rnorm(length(set.k), 1)
}

Y <- apply(y, 2, scale) * sqrt(h2) + .rnorm(n, m) * sqrt(1 - h2)
```

After standardization, we can easily compute the z-score matrix.

```{r}
X <- apply(X, 2, scale)
Y <- apply(Y, 2, scale)
Z <- crossprod(X, Y) / sqrt(nrow(X))
```

```{r echo=FALSE, message=FALSE}
library(tidyverse)
theme_set(theme_classic())

z.melt <- reshape2::melt(Z, value.name = "z") %>%
    as_tibble()
```

```{r fig.width=8, fig.height=5}
plt <-
    ggplot(z.melt, aes(x = Var1, y = -2 * pnorm(abs(z), lower.tail=FALSE, log.p=TRUE)/log(10))) +
    xlab("variants") + ylab("-log10 p-value") +
    facet_wrap(.~Var2, nrow = 2) +
    geom_point(alpha=.8, colour="gray") +
    geom_vline(xintercept = common, lty=2, color="green")

.melt <- z.melt %>% filter(`Var1` %in% common)
plt <- plt + geom_point(data = .melt, color="red", pch=21, stroke=1)
.melt <- z.melt %>% filter(`Var1` %in% other)
plt <- plt + geom_point(data = .melt, color="blue", pch=22, stroke=1)
print(plt)
```

Here, we marked the shared causal variant with the red circles and trait-specific ones with the blue squares.

### Why don't we run `susieR` independently for each trait to find credible sets?

```{r}
susie.result <- NULL
for(l in 1:ncol(Y)){
    susie.l <- susieR::susie(X, Y[,l])
    susie.result <- rbind(susie.result, data.frame(Var1 = 1:ncol(X), Var2 = l, pip = susie.l$pip))
}
```

```{r}
plt <-
    ggplot(susie.result, aes(x = Var1, y = pip)) +
    xlab("variants") + ylab("PIP") +
    facet_wrap(.~Var2, nrow = 2) +
    geom_point(alpha=.8, colour="gray") +
    geom_vline(xintercept = common, lty=2, color="green")

.melt <- susie.result %>% filter(`Var1` %in% common)
plt <- plt + geom_point(data = .melt, color="red", pch=21, stroke=1)
.melt <- susie.result %>% filter(`Var1` %in% other)
plt <- plt + geom_point(data = .melt, color="blue", pch=22, stroke=1)
print(plt)
```

Okay, it works well. However, it doesn't take full advantage of a multi-trait regression setting. First, we can see that the shared variants are dropped in many traits. Moreover, when the shared one is selected, the other trait-specific effect is often not selected. Can we identify all the variants that may have contributed to any of the traits?

### Can we leverage the information shared across traits?

Here comes our `mtSusie` (multi-trait Sum of Single Effects) model. The idea is a simple modification to the original `susieR` method. For each level/layer, we aggregate the log Bayes Factors across multiple traits. Yes, it was that simple. 

```{r}
library(mtSusie)
mt.susie <- mt_susie(X, Y, clamp=8)
```

```{r fig.width=8, fig.height=5}
mt.susie.dt <- data.table::setDT(mt.susie$cs)

plt <- ggplot(mt.susie.dt, aes(x = variants, y = alpha)) +
    xlab("variants") + ylab("PIP") +
    facet_wrap(.~traits, nrow = 2) +
    geom_point(alpha=.8, colour="gray") +
    geom_vline(xintercept = common, lty=2, color="green")    

.melt <- mt.susie.dt %>% filter(`variants` %in% common)
plt <- plt + geom_point(data = .melt, color="red", pch=21, stroke=1)
.melt <- mt.susie.dt %>% filter(`variants` %in% other)
plt <- plt + geom_point(data = .melt, color="blue", pch=22, stroke=1)
print(plt)
```

Once we have identified a "union" of all the causal variants across all the traits, we can assign trait-specific effects based on variant-by-variant statistics at each level/layer. At each level, `mtSusie` will select a causal variant shared across traits, but not all the traits will support it and result in trait-specific local false sign rate (LFSR) values.

```{r fig.width=8, fig.height=3.5}
.show <- unique(mt.susie.dt[alpha > .1]$variants)
.dt <- mt.susie.dt[variants %in% .show]

ggplot(.dt, aes(x=as.factor(variants), y=as.factor(traits))) +
    facet_wrap(.~`levels`, scales="free", nrow=1) +
    xlab("variant") + ylab("trait") +
    geom_point(aes(fill=`z`, size=-log10(pmax(`lfsr`,1e-3))), pch=22) +
    scale_size_continuous(range = c(0,5), "LFSR", labels=function(x) round(10^(-x), 3)) +
    scale_fill_gradient2("z-score", mid="white", high="red", low="blue")
```

